EDIpack: A generic and interoperable Lanczos-based solver for quantum impurity problems

### 2025-05-21 05_examples.tex
Fixing EDIpack and packages names in the paper: EDIpack EDIpack2ineq ----> libedipack.a EDIpack_C ----> libedipack_cbindings.so EDIpack2py EDIpack2jl EDIpack2Triqs EDIpack-Conda
Co-authored-by: Adriano Amaricci <adriano.amaricci@gmail.com>

\setbox0=\hbox{%
  \begin{minipage}{0.465\linewidth}
    \center{\bf Fortran}
\begin{lstlisting}[style=fstyle,frame=none,numbers=none,basicstyle={\scriptsize\ttfamily}]
program ed_hm_bethe
   USE EDIPACK2
   USE SCIFOR
   implicit none
   integer                :: Le=1000
   real(8)                :: wmixing=0.5d0
   real(8)                :: D=1d0
   integer                :: Nb
   real(8),allocatable    :: Bath(:)
   complex(8),allocatable :: Hloc(:,:,:,:)
   real(8),allocatable    :: Ebands(:),Dbands(:)
   complex(8),allocatable :: Smats(:,:,:,:,:)
   complex(8),allocatable :: Delta(:,:,:,:,:)
   ...  
   !Read variables
   call ed_read_input('inputED.conf')
   !Solver-specific arays
   allocate(Smats(Nspin,Nspin,Norb,Norb,Lmats))
   allocate(Delta(Nspin,Nspin,Norb,Norb,Lmats))
   allocate(Hloc(Nspin,Nspin,Norb,Norb));Hloc=0d0
   ...
   !Initialize Bethe arrays
   allocate(Ebands(Le),Dbands(Le))
   Ebands = linspace(-D,D,Le,mesh=de)
   Dbands = dens_bethe(Ebands,Wband)*de
   !Set local H
   call ed_set_hloc(Hloc)
\end{lstlisting}
\end{minipage}
}
\savestack{\listingA}{\box0}

\setbox0=\hbox{%
  \begin{minipage}{0.505\linewidth}
    \center{\bf C++}
\begin{lstlisting}[style=cstyle,frame=none,numbers=none,basicstyle={\scriptsize\ttfamily}]
#include <edipack2_cbinding.h>
using namespace std;
...
//Main variables
int Le = 1000;
int iloop = 0;
double wmixing = 0.5;
double D = 1.0;
//Read  variables    
char input[] = "inputED.conf"; 
read_input(input);      
//Dimensions
int64_t d[4] = {Nspin,Nspin,Norb,Norb};
int total_size = d[0] * d[1] * d[2] * d[3];
int total_size_n5 = total_size * Lmats;    
// Solver-specific arrays
vector<complex<double>> Hloc(total_size);
vector<complex<double>> Smats(total_size_n5);
vector<complex<double>> Delta(total_size_n5);
...
// Initialize Bethe arrays
vector<double> Ebands, Dbands;
Ebands = linspace(-D,D,Le);
Dbands = dens_bethe(Ebands,Wband,de);
//set local H
ed_set_Hloc_single_N4(Hloc.data(), d);
\end{lstlisting}
\end{minipage}
}
\savestack{\listingB}{\box0}

@@

\begin{figure}%[t!]
  \includegraphics[width=\linewidth]{figures/figBetheW2D.pdf}
    \caption{\label{figEx1W}%
        }
\end{figure}

\subsubsection{Finite temperature solution (W2Dynamics interface)}\label{SecExamplesBetheDMFTw2d}
To demonstrate use of the w2dynamics interface, here we briefly
discuss how to solve the DMFT problem for the Bethe lattice using
W2Dynamics interface. \fixme{For definitiness we compare the finite low
temperature results obtained via the two solvers: ED and CTQMC.  }

Unlike when the Fortran API is used, none of the
five steps of the iterative algorithm listed there are left to the
user to implement themselves. Instead, in addition to possibly
non-default values for \NAME input variables, the model specification
must also be added to a w2dynamics parameter file and depending on the
model, further input files might need to be provided. Further
differences are that initialization is performed in w2dynamics by
starting from a self-energy of zero or reading a self-energy value
from an old file, which results in the initial Weiss field that the
initial bath parameters are fitted to, and that automatic convergence
checking is not currently available. Finally, using the same settings
as for the Fortran example we arrive at the w2dynamics parameter file
shown in the following listing.

\begin{lstlisting}[style=fstyle,language={},numbers=none,basicstyle={\scriptsize\ttfamily}]
[General]
DOS             = Bethe           # support for the Bethe lattice is built-in
half-bandwidth  = 1               # list of half-bandwidths per orbital
NAt             = 1               # number of impurities
beta            = 1000            # inverse temperature, defines Matsubara grid spacing
mu              = 1.0             # chemical potential set to achieve half-filling
EPSN            = 0.0             # turns off filling-based chemical potential search
DMFTsteps       = 100             # given no convergence checking, we might want fewer
magnetism       = para            # symmetrize self-energies per spin
FileNamePrefix  = bethe_dmft_U2   # prefix for the output file name
fileold         = bethe_dmft*hdf5 # file to read an initial self-energy from
readold         = 0               # iteration number to read initial self-energy from, 0 turns off
mixing          = 0.5             # mixing, but mixes self-energies and not Weiss fields
mixing_strategy = linear          # linear mixing as in the Fortran example
solver          = EDIPACK         # use EDIpack2 as impurity solver, not default CTHYB

@@

[QMC]                             # Parameters for some grid sizes are part of this section
Ntau            = 1024            # imaginary time grid size
Niw             = 4096            # number of positive Matsubara frequencies

[EDIPACK]                         # further non-default parameters, also taken from Fortran example
NBATH           = 7               # number of bath sites
ED_TWIN         = True            # use twin symmetry
LFIT            = 2048            # number of Matsubara frequencies used for the bath fit
LANC_NGFITER    = 500             # number of Lanczos iterations for Green's function
CG_FTOL         = 1e-10           # conjugate-gradient tolerance
CG_NITER        = 2048            # maximum number of conjugate-gradient iterations
\end{lstlisting}

With this parameter file, we can use the {\tt DMFT.py} script of
w2dynamics to perform the execution and find the results in form of
both a typical w2dynamics HDF5 output file whose name starts with {\tt
  bethe\textunderscore{}dmft\textunderscore{}U2} and a number of folders whose names start with the
name of the HDF5 output file that contain unavoidable or explicitly
requested \NAME input and output files for each individual DMFT
iteration. We can extract results from the HDF5 output file e.g. using
the {\tt hgrep} script of w2dynamics. The following listing shows how
to perform a calculation with 10 parallel MPI processes and print the
impurity self-energy result of the last DMFT iteration to the terminal
assuming that the parameter file is {\tt bethe\textunderscore{}dmft.in} in the working
directory and w2dynamics installed to {\tt /opt/w2dynamics}.

\begin{lstlisting}[style=mybash]
mpiexec -n 10 /opt/w2dynamics/DMFT.py bethe_dmft.in
python /opt/w2dynamics/hgrep latest siw-full -1
\end{lstlisting}

@@

%
The main effect of the interaction is enclosed in the renormalized
mass term $M_{eff}=M+\tfrac{1}{4}\Tr{\Re\Sigma(i\omega\to0)}$. The
real-part of the self-energy being proportional to $\Gamma_{03}$
corrects the mass term with respect to its bare value. To leading
order this correction is given proportional to $\tfrac{U-5J}{4}\langle
\hat{T}_z\rangle$. So, for our choice of parameters, the effect of
interaction would be to reduce the (effective) mass term and thus
reducing the topological band gap. In the strong-coupling limit the
two orbitals get populated with one electron per site, reaching the
conditions for the formation of a Mott insulating state.
As reported in panel (A) of \figu{figEx4}, at intermediate coupling,
the DMFT solutions features the formation of a
region of exciton condensation with $\langle E_x\rangle>0$, $\langle
E_0\rangle=\langle E_z\rangle=0$. 
Unlike the static mean-field description~\cite{Blason2021}, the
transition from the QSHI topological state to the Excitonic Insulator 
(EI) becomes of first-order in presence of quantum fluctuations
corrections~\cite{Continentino,Paoletti} contained in DMFT. On the
other hand, the EI continuously evolves into a Mott Insulator (MI)
for larger interaction (neglecting for simplicity any 
anti-ferromagnetic ordering that would naturally occurr in this
regime).
In the same panel we report the evolution of 
$\langle T_z\rangle$ to illustrate the progressive reduction of the
orbital polarization and, thus, of the effective mass.

>>>>

\setbox0=\hbox{%
  \begin{minipage}{0.465\linewidth}
    \center{\bf Fortran}
\begin{lstlisting}[style=fstyle,frame=none,numbers=none,basicstyle={\scriptsize\ttfamily}]
program ed_hm_bethe
   USE EDIPACK
   USE SCIFOR
   implicit none
   integer                :: Le=1000
   real(8)                :: wmixing=0.5d0
   real(8)                :: D=1d0
   integer                :: Nb
   real(8),allocatable    :: Bath(:)
   complex(8),allocatable :: Hloc(:,:,:,:)
   real(8),allocatable    :: Ebands(:),Dbands(:)
   complex(8),allocatable :: Smats(:,:,:,:,:)
   complex(8),allocatable :: Delta(:,:,:,:,:)
   ...  
   !Read variables
   call ed_read_input('inputED.conf')
   !Solver-specific arays
   allocate(Smats(Nspin,Nspin,Norb,Norb,Lmats))
   allocate(Delta(Nspin,Nspin,Norb,Norb,Lmats))
   allocate(Hloc(Nspin,Nspin,Norb,Norb));Hloc=0d0
   ...
   !Initialize Bethe arrays
   allocate(Ebands(Le),Dbands(Le))
   Ebands = linspace(-D,D,Le,mesh=de)
   Dbands = dens_bethe(Ebands,Wband)*de
   !Set local H
   call ed_set_hloc(Hloc)
\end{lstlisting}
\end{minipage}
}
\savestack{\listingA}{\box0}

\setbox0=\hbox{%
  \begin{minipage}{0.505\linewidth}
    \center{\bf C++}
\begin{lstlisting}[style=cstyle,frame=none,numbers=none,basicstyle={\scriptsize\ttfamily}]
#include <edipack_cbindings.h>
using namespace std;
...
//Main variables
int Le = 1000;
int iloop = 0;
double wmixing = 0.5;
double D = 1.0;
//Read  variables    
char input[] = "inputED.conf"; 
read_input(input);      
//Dimensions
int64_t d[4] = {Nspin,Nspin,Norb,Norb};
int total_size = d[0] * d[1] * d[2] * d[3];
int total_size_n5 = total_size * Lmats;    
// Solver-specific arrays
vector<complex<double>> Hloc(total_size);
vector<complex<double>> Smats(total_size_n5);
vector<complex<double>> Delta(total_size_n5);
...
// Initialize Bethe arrays
vector<double> Ebands, Dbands;
Ebands = linspace(-D,D,Le);
Dbands = dens_bethe(Ebands,Wband,de);
//set local H
ed_set_Hloc_single_N4(Hloc.data(), d);
\end{lstlisting}
\end{minipage}
}
\savestack{\listingB}{\box0}

@@

\subsubsection{Finite temperature solution (w2dynamics interface)}\label{SecExamplesBetheDMFTw2d}
To demonstrate the usage of the w2dynamics interface to \NAME,  here we briefly discuss how to solve the same problem, i.e. the Hubbard model on the Bethe lattice within DMFT using w2dynamics. 
In order to showcase the capabilities of \NAME to address low-temperature problems we compare the continuous-time Quantum Monte Carlo (CTQMC) solver using the hybridization expansion method (CT-HYB) included in w2dynamics against the \NAME solver at finite temperature.

In contrast to the \NAME API that only enables access to the impurity solver and bath optimization procedures, but requires the user to implement (using different methods) the DMFT algorithm themselves, the standard way to use w2dynamics is fundamentally different: a single Python script {\tt DMFT.py} is provided that takes care of the whole DMFT calculation via dedicated classes to implement generic self-consistency. The w2dynamics calculation is then entirely controlled by a model-dependent parameters file {\tt Parameters.in}, which contains a number of variable specifications including options to control the ED solver inherited from \NAME. Further information about the functioning of w2dynamics can be found in Ref.~\onlinecite{Wallerberger2019CPC}

Another important difference concerns the initial point of the iterative DMFT solution algorithm: while \NAME starts from a given discrete bath, w2dynamics is initialized with a zero self-energy function or alternatively reads this quantity from a converged solution file. When using the \NAME ED interface, the initial Weiss field is determined using self-consistency and finally an initial discrete bath is obtained through the \NAME bath optimization procedure.

The w2dynamics configuration file for the solution of the Bethe lattice problem at finite temperature reads: 
\begin{lstlisting}[style=mybash,language={},numbers=none,basicstyle={\scriptsize\ttfamily}]
[General]
DOS             = Bethe           # support for the Bethe lattice is built-in
half-bandwidth  = 1               # list of half-bandwidths per orbital
NAt             = 1               # number of impurities
beta            = 100             # inverse temperature
mu              = 1.0             # chemical potential set to achieve half-filling
EPSN            = 0.0             # turns off filling-based chemical potential search
DMFTsteps       = 100             # given no convergence checking, we might want fewer
magnetism       = para            # symmetrize self-energies per spin
FileNamePrefix  = bethe_dmft_U2   # prefix for the output file name
fileold         = bethe_dmft*hdf5 # file to read an initial self-energy from
readold         = 0               # iteration number to read initial self-energy from, 0 turns off
mixing          = 0.5             # mixing, but mixes self-energies and not Weiss fields
mixing_strategy = linear          # linear mixing as in the Fortran example
FTType          = none            # (for CT-HYB): use the NFFT-measured G
solver          = EDIPACK         # use EDIpack as impurity solver, not default CTHYB

@@

[EDIPACK]                         # further ED parameters, as in the Fortran example
NBATH           = 7               # number of bath sites
ED_TWIN         = True            # use twin symmetry
LFIT            = 2048            # number of Matsubara frequencies used for the bath fit
LANC_NGFITER    = 500             # number of Lanczos iterations for Green's function
CG_FTOL         = 1e-10           # conjugate-gradient tolerance
CG_NITER        = 2048            # maximum number of conjugate-gradient iterations
ED_FINITE_TEMP  = True            # at finite temperature T = 1/beta

[QMC]                             # Parameters for some grid sizes and the CT-HYB calculation
Ntau            = 1024            # imaginary time grid size
Niw             = 4096            # number of positive Matsubara frequencies
# parameters only relevant for CT-HYB follow
MeasGiw         = 1               # enable NFFT measurement of G
NCorr           = 175             # estimate of the autocorrelation length
Nmeas           = 200000          # number of measurements / sample size
Nwarmups        = 1000000         # number of initial warmup steps for Markov chain thermalization
\end{lstlisting}
Assuming that w2dynamics is installed to {\tt /opt/w2dynamics} and the preceding parameters file saved as {\tt bethe\textunderscore{}dmft.in}, we can start the calculation by running the {\tt DMFT.py} script:
\begin{lstlisting}[style=mybash,numbers=none]
mpiexec -n 10 /opt/w2dynamics/DMFT.py bethe_dmft.in
\end{lstlisting}

The results are collected into an HDF5 archive in the usual w2dynamics format, including output quantities inherited from \NAME. Results can be viewed using the script {\tt hgrep} provided with w2dynamics or with other HDF5 tools. In this example we extract the Matsubara self-energy function $\Sigma(i\omega_n)$ for the last DMFT iteration:
\begin{lstlisting}[style=mybash,numbers=none]
/opt/w2dynamics/hgrep latest siw-full -1
\end{lstlisting}

\begin{figure}%[t!]
  \includegraphics[width=\linewidth]{figures/figBetheW2D.pdf}
    \caption{\label{figEx1W}%
    \textbf{Finite temperature DMFT solution.}
    Comparison of the imaginary parts of the Green's function $\Im{G}(i\omega_n)$ and Self-energy $\Im{\Sigma}(i\omega_n)$ from different solutions of the Hubbard model on the Bethe lattice using DMFT. Data are for $T/D=0.01$ and $U/D=2.0$. CTQMC results from the solver included with w2dynamics are compared against \NAME ED results, used both through the w2dynamics interface and with a standalone Fortran program. 
        }
\end{figure}

In \figu{figEx1W} we report a comparison of the results obtained using the hybridization expansion CTQMC solver in w2dynamics and using the \NAME ED method both as a solver in w2dynamics and with a program using the Fortran API directly as shown in the previous subsection. Using DMFT, we solve the Hubbard model on a Bethe lattice for $T/D=0.01$ and $U/D=2.00$, corresponding to a correlated metallic state. 
In panel (A) we show the behavior of the imaginary part of the impurity Green's function $\Im{G}(i\omega_n)$ (note that w2dynamics does not directly have access to the real-axis Green's function when using the CTQMC solver). In panel (B) we show the same comparison for impurity self-energy $\Im{\Sigma}(i\omega_n)$. 
The results obtained from the three calculations are in this case nearly identical already for a small number of bath levels ({\tt NBATH=7}) and relatively low QMC statistics.

@@

%
The main effect of the interaction is enclosed in the renormalized
mass term $M_{eff}=M+\tfrac{1}{4}\Tr{\Re\Sigma(i\omega\to0)}$. The
real-part of the self-energy being proportional to $\Gamma_{03}$
corrects the mass term with respect to its bare value. To leading
order this correction is given proportional to $\tfrac{U-5J}{4}\langle
\hat{T}_z\rangle$. So, for our choice of parameters, the effect of
interaction would be to reduce the (effective) mass term and thus
reducing the topological band gap. In the strong-coupling limit the
two orbitals get populated with one electron per site, reaching the
conditions for the formation of a Mott insulating state.
As reported in panel (A) of \figu{figEx4}, at intermediate coupling,
the DMFT solutions features the formation of a
region of exciton condensation with $\langle E_x\rangle>0$, $\langle
E_0\rangle=\langle E_z\rangle=0$. 
Unlike the static mean-field description~\cite{Blason}, the
transition from the QSHI topological state to the Excitonic Insulator 
(EI) becomes of first-order in presence of quantum fluctuations
corrections~\cite{Continentino,Paoletti} contained in DMFT. On the
other hand, the EI continuously evolves into a Mott Insulator (MI)
for larger interaction (neglecting for simplicity any 
anti-ferromagnetic ordering that would naturally occurr in this
regime).
In the same panel we report the evolution of 
$\langle T_z\rangle$ to illustrate the progressive reduction of the
orbital polarization and, thus, of the effective mass.

### 2025-05-21 edipack2.tex
Fixing EDIpack and packages names in the paper: EDIpack EDIpack2ineq ----> libedipack.a EDIpack_C ----> libedipack_cbindings.so EDIpack2py EDIpack2jl EDIpack2Triqs EDIpack-Conda
Co-authored-by: Adriano Amaricci <adriano.amaricci@gmail.com>

\title{EDIpack 2.0: A generic and interoperable Lanczos-based solver for quantum impurity problems}
\author[a,b,c]{L.~Crippa*}
\author[a]{I.~Krivenko*}
\author[d]{S.~Giuli*}
\author[d]{G.~Bellomia*}
\author[c]{A.~Kowalski}
\author[e]{F.~Petocchi}
\author[f]{M.~Wallerberger}
\author[g]{G.~Mazza}
\author[h]{L.~de Medici}
\author[c,b]{G.~Sangiovanni}
\author[d,i]{M.~Capone*}
\author[i]{A.~Amaricci*}
% \author[]{A.~Scazzola}

@@

\section{Conclusions}
{\color{red} Tentative conclusions}
We have presented EDIpack2, a versatile, high-performance solver for
generic quantum impurity problems. Leveraging on the massively
parallel algorithms introduced with its predecessor, this version of
the library features new capabilities and symmetries within a unified
framework, providing reliable evaluations of the dynamical correlation
functions on arbitrary complex argument.
This makes it particularly well-suited for capturing the sharp
low-energy excitations and complex spectral features of strongly
correlated systems. We have shown that this new version enables the evaluation of impurity
reduced density matrices directly from Fock space quantities and thus unlock 
quantum information analyses of correlated systems.

A central feature of EDIpack2 is its strong focus on interoperability,
achieved through modern Fortran constructs, C/C++-bindings, and
comprehensive APIs for Python (EDIpy2). 
These interfaces enable seamless integration with broader
computational frameworks like TRIQS and W2Dynamics, expanding the
functionality of these platforms and providing a robust foundation for
reproducible research.

We thoroughly discussed the implementation of the EDIpack2 library, the
most important algorithms and classes. We presented in
details the third-party interfaces which extend exact diagonalization
capabilities beyond the realm of the library itself.   
Finally, we showcased the use of the EDIpack2 software in different
contexts via elaborated example which can serve as reference for the
potential users.

The modular and extensible design of EDIpack2 provides a
natural foundation for future extensions, for instance to cluster-DMFT where the
impurity problem is generalized to include clusters of interacting
sites. This approach would enable the treatment of spatial
correlations beyond the single-site approximation, capturing effects
such as d-wave pairing in high-temperature superconductors, charge
order, and complex spin textures in strongly correlated systems.
We anticipate that EDIpack2 will become a valuable tool for the
computational condensed matter community, supporting a wide range of
studies on strongly correlated materials and providing a reliable
reference platform for quantum impurity solvers.

>>>>

\title{EDIpack: A generic and interoperable Lanczos-based solver for quantum impurity problems}
\author[a,b,c]{L.~Crippa*}
\author[a]{I.~Krivenko*}
\author[d]{S.~Giuli*}
\author[d]{G.~Bellomia*}
\author[c]{A.~Kowalski}
\author[e]{F.~Petocchi}
\author[f]{M.~Wallerberger}
\author[g]{G.~Mazza}
\author[h]{L.~de Medici}
\author[c,b]{G.~Sangiovanni}
\author[d,i]{M.~Capone*}
\author[i]{A.~Amaricci*}
% \author[]{A.~Scazzola}

@@

\section{Conclusions}
{\color{red} Tentative conclusions}
We have presented \NAME{}, a versatile, high-performance solver for
generic quantum impurity problems. Leveraging on the massively
parallel algorithms introduced with its predecessor, this version of
the library features new capabilities and symmetries within a unified
framework, providing reliable evaluations of the dynamical correlation
functions on arbitrary complex argument.
This makes it particularly well-suited for capturing the sharp
low-energy excitations and complex spectral features of strongly
correlated systems. We have shown that this new version enables the evaluation of impurity
reduced density matrices directly from Fock space quantities and thus unlock 
quantum information analyses of correlated systems.

A central feature of \NAME is its strong focus on interoperability,
achieved through modern Fortran constructs, C/C++-bindings, and
comprehensive APIs for Python (EDIpy2). 
These interfaces enable seamless integration with broader
computational frameworks like TRIQS and W2Dynamics, expanding the
functionality of these platforms and providing a robust foundation for
reproducible research.

We thoroughly discussed the implementation of the \NAME library, the
most important algorithms and classes. We presented in
details the third-party interfaces which extend exact diagonalization
capabilities beyond the realm of the library itself.   
Finally, we showcased the use of the \NAME software in different
contexts via elaborated example which can serve as reference for the
potential users.

The modular and extensible design of \NAME provides a
natural foundation for future extensions, for instance to cluster-DMFT where the
impurity problem is generalized to include clusters of interacting
sites. This approach would enable the treatment of spatial
correlations beyond the single-site approximation, capturing effects
such as d-wave pairing in high-temperature superconductors, charge
order, and complex spin textures in strongly correlated systems.
We anticipate that \NAME will become a valuable tool for the
computational condensed matter community, supporting a wide range of
studies on strongly correlated materials and providing a reliable
reference platform for quantum impurity solvers.
